---
title: "Dispersion"
subtitle: ": Lecture 4"
author: "Dave Brocker"
institute: "Farmingdale State College"
footer: "⬡⬢⬡⬢"
format: 
  revealjs:
    theme: custom.scss
    incremental: true   
    touch: true
    scrollable: true
    chalkboard: true
---

## Descriptive Statistics

X = refers to the value for 1 person (1 datapoint) in the sample

$\bar{x}$ = refers to the mean or average value of X in the sample

$\sum(x)$ = refers to the sum of all x values (sum of all the datapoints)

$x^2$ = refers to the squared value of x (x multiplied by x)

## Descriptive Statistics

### Describe the characteristics of a sample in terms of:

-   Central tendency and

-   Dispersion (aka Variability aka Variance)

## Central tendency

### What is it, when should we use it, and how do we calculate it?

-   Mean

-   Median

-   Mode

## Central tendency

### What do measures of Central Tendency tell you?

Central tendency

What do measures of Central Tendency tell you?

-   The mid-point in the data

-   The answer most of the participants gave.

## Measures of Dispersion

### What does dispersion mean?

-   Spread-out-ness

-   How much the participants' responses differ from one another

-   Variance

## Measures of Dispersion

### Measures of dispersion tell us about the spread-out-ness of the data, but what does that *actually* mean?

## Measures of Dispersion

### Measures of dispersion tell us about the spread-out-ness of the data, but what does that *actually* mean?

-   If I am measuring HYPE, what would low dispersion tell me?

-   If I am measuring HYPE, what would high dispersion tell me?

## Measures of Dispersion

### Measures of dispersion tell us about consensus.

-   Did participants give similar answers?

-   Did participants give wildly different answers?

## Dispersion

-   Data should have natural dispersion.

-   If everyone gives a similar answer, it's harder to analyze difference.

## Calculating Dispersion

### How do you think we should calculate dispersion?

-   ON AVERAGE, how far is each X-VALUE from the MIDPOINT.

    -   Calculate how far each X-VALUE is from the MIDPOINT.

    -   Take the AVERAGE of those distances.

## Calculating Dispersion

In psychology, the mid-point we use will be the [***mean***]{.underline}.

## Calculating Dispersion

### How do you think we should calculate dispersion?

-   ON AVERAGE, how far is each X-VALUE from the MEAN.

    -   **Calculate how far each X-VALUE is from the MIDPOINT.** [^1]

    -   Take the AVERAGE of those distances.

[^1]: Deviation Score

## Deviation Scores

### Calculate how far each X-VALUE is from the MEAN.

$$
x - \bar{x}
$$

## Calculating Dispersion

### Step 1: Calculate the mean.

```{r}
library(huxtable)

data.frame(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2)
) |> hux() |> 
  theme_article()
```

$$
\frac{\sum(x)}{n}
$$

$$X = 20 / 10 = 2$$

## Calculating Dispersion

### Step 2: Subtract the mean from each X-value.

```{r}
library(dplyr)
tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X)
) |> hux() |> 
  theme_article()

```

-   The average of the deviation scores will always = 0.

-   On average, the x-values = x.

-   So the average distance of x from x is 0.

## Calculating Dispersion

### How do you think we should calculate dispersion?

1.  Calculate the mean.
2.  Calculate the deviation scores.
3.  What can we do to the deviation scores to prevent them from adding up to 0?
    1.  \[Make this pop out\]
4.  Take the AVERAGE

## What is the Mean of the Deviation Scores?

-   The problem is the negative numbers!

-   How do we get rid of negative numbers?

```{r}
tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) |> 
  # Bold Last Column
  hux() |> 
  theme_article()
```

## Calculating Dispersion

-   Step 1: Calculate the mean.

-   

    ```         
      X = 20 / 10 = 2
    ```

-   Step 2: Subtract the mean from each X-value.

-   Step 3: Square each deviation score.

## Calculating Dispersion

### Now what: Calculate the average squared deviation.

-   Step 4: Add up the Squared Deviations.

```{r}
le4 <- 
  tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) 

le4


le4$`(x-M)^2` |> sum()
```

## Calculating Dispersion

-   Step 5: Divide by (n-1).

\[insert Cat? GIF\]

-   (n-1) refers to the degrees of freedom... don't worry about it for now.

```{r}
le4 <- 
  tibble(
  ParticipantID = 1:10,
  X = c(1,3,2,2,3,2,1,3,1,2),
  `x-M` = X-mean(X),
  `(x-M)^2` = `x-M`^2
  ) 

sum(le4$`(x-M)^2`)/
  length((le4$`(x-M)^2`)-1)
```

### Variance \^

## Calculating Dispersion

### How do you think we should calculate dispersion?

-   ON AVERAGE, how far is each X-VALUE from the MEAN.

-   **Calculate how far each X-VALUE is from the MIDPOINT.**

    -   \[Make pop-out we did it!...right?\]

-   Take the AVERAGE of those distances.

## Calculating Dispersion

### How do think we should calculate dispersion?

-   When we squared the deviation scores in Step 3, we inflated the deviation.

    -   We have to undo that inflation.

    -   How do you undo squaring?

    -   We take the square root.

## Calculating Dispersion

-   Step 6: Take the square root.

$$
\sqrt{.667} = .816
$$

## Calculating standard deviation

1.  Mean: Calculate the mean.
2.  Deviation scores: Subtract the mean from each x-value.
3.  Squared deviations: Square each deviation score.
4.  Sum of squares: Add up the squared deviations.
5.  Variance: Divide the sum of squares by (n-1).
6.  Standard Deviation: Take the square root of the variance.

## Formulas for Dispersion

### Variance

$$
s^2 = \frac{\sum(x-\bar{x})^2}{N-1}
$$

## Formulas for Dispersion

### Standard Deviation

$$
s = \sqrt{\frac{\sum(x-\bar{x})^2}{N-1}}
$$

-   Variance is a measure of dispersion.

    -   It's not helpful, because it's inflated (from squaring the deviation scores).

-   Standard deviation is a better measure of dispersion, because it is standardized.

    -   This means that a standard deviation of 1 means a distance of 1 on the scale used to measure x.

    -   HYPE scale: 1 to 10: standard deviation of 1 means the distance from one rating to the next, from 1 to 2.

    -   If the mean = 5, SD = 1, most scores fall between 4 and 6.

## Variance & Standard deviation

-   The Variance is the standard deviation squared: s²

-   The Standard Deviation is the square root of the variance: s
